Fast Sorting on Modern Computers

### 2005-05-31 paper/docs/background.tex
mostly removed the use of the word project, and 'I'

The ADT interface in this project is taken from \cite{Sedgewick02}. An \cc{Item}
is a record, and a function is defined to extract its key.  Comparisons are done
via the \cc{less} function, and swapping is done using the \cc{exch} function.
These are replaced with their actual functions by the C preprocessor. The code
used for ADTs is in Figure \vref{Definitions used by the ADT}. The tests
presented in this report use \cc{unsigned integer}s as Items, which were
compared using a simple \n{less-than} function\footnote{This abstraction broke
down in several cases, such as where a \n{greater-than} function was required,
i.e. in quicksort.}.  In this case, a record and key are equivalent, and the
word \n{key} is used to refer to items to be sorted.

@@

The address used by the cache is not necessarily the address used by a program.
The addresses used within a program 
are called \n{virtual addresses}. Some caches are addressed by virtual address,
some caches by \n{physical address}, and some caches by \n{bus
address}\footnote{A bus address is used to address parts of the computer
connected over a bus, including main memory, multimedia controllers such as
sound and graphics cards, and storage devices such as hard drivers and floppy
disk controllers.}. Several optimizations in later chapters use the address to
offset arrays so that conflicts are not caused. However, because a program
cannot be sure of what type of addressing is used, the actual manipulated
address may not directly relate to the index used to address the cache. This
problem is not encountered in simulations in this project, as SimpleScalar
addresses its caches by virtual address.

Caches work on the principle of \n{locality}. It is expected that after some
data has been used once, it is likely to be used again in the near future. This
is called \n{\label{temporal locality}temporal locality}. Similarly, if some data is being used, it is
likely that data near it will also be used soon, such as two consecutive
elements of an array. This is called \n{spatial locality}. Many of the
algorithms in this project are designed to take advantage of locality. Spatial
locality is exploited by ensuring that once a cache line is loaded, each key in
the cache line is used. Temporal locality is exploited by ensuring that an
algorithm uses data as much as it can as soon as it is loaded, rather than over
a large time scale, during which it may have been ejected from the cache.

@@

For this project, I split branches into two types. The first is a \n{flow
control branch}. Flow control refers to controlling the flow of execution of a
program, and branches of this type are loops and other control statements.
The second type is a \n{comparative branch}: these are comparisons between two
pieces of data, such as two keys in an array being sorted. A branch can be both
a flow control and a comparative branch, but these will be referred to as
comparative branches here. This is due to the properties of the branches: a
comparative branch should not be easily predictable; a flow control branch
should be easy to predict based on previous predictions of the same branch.

There are several kinds of branch predictors: \n{static}, \n{semi-static} and
\n{dynamic}. This project is only interested in dynamic predictors, due to
their high levels of accuracy. The other types are none-the-less discussed
below, as an introduction to dynamic predictors. A good discussion on predictor
types is \cite{Uht97}, from which the percentages below are taken. These
predictors were tested by Uht on the SPECint92 benchmarking suite. This is a
standard suite of integer-based programs used to measure the performance of an
architecture against another. Programs in this benchmark include \cc{gcc}, a
major C compiler, \cc{espresso}, which generates Programmable Logic Arrays and
\cc{li}, a LISP interpreter.

>>>>

The ADT interface used here is taken from \cite{Sedgewick02}. An \cc{Item}
is a record, and a function is defined to extract its key.  Comparisons are done
via the \cc{less} function, and swapping is done using the \cc{exch} function.
These are replaced with their actual functions by the C preprocessor. The code
used for ADTs is in Figure \vref{Definitions used by the ADT}. The tests
presented in this report use \cc{unsigned integer}s as Items, which were
compared using a simple \n{less-than} function\footnote{This abstraction broke
down in several cases, such as where a \n{greater-than} function was required,
i.e. in quicksort.}.  In this case, a record and key are equivalent, and the
word \n{key} is used to refer to items to be sorted.

@@

The address used by the cache is not necessarily the address used by a program.
The addresses used within a program 
are called \n{virtual addresses}. Some caches are addressed by virtual address,
some caches by \n{physical address}, and some caches by \n{bus
address}\footnote{A bus address is used to address parts of the computer
connected over a bus, including main memory, multimedia controllers such as
sound and graphics cards, and storage devices such as hard drivers and floppy
disk controllers.}. Several optimizations in later chapters use the address to
offset arrays so that conflicts are not caused. However, because a program
cannot be sure of what type of addressing is used, the actual manipulated
address may not directly relate to the index used to address the cache. This
problem is not encountered in our simulations, as SimpleScalar
addresses its caches by virtual address.

Caches work on the principle of \n{locality}. It is expected that after some
data has been used once, it is likely to be used again in the near future. This
is called \n{\label{temporal locality}temporal locality}. Similarly, if some data is being used, it is
likely that data near it will also be used soon, such as two consecutive
elements of an array. This is called \n{spatial locality}. Many of the
algorithms presented here are designed to take advantage of locality. Spatial
locality is exploited by ensuring that once a cache line is loaded, each key in
the cache line is used. Temporal locality is exploited by ensuring that an
algorithm uses data as much as it can as soon as it is loaded, rather than over
a large time scale, during which it may have been ejected from the cache.

@@

In this report, we split branches into two types. The first is a \n{flow
control branch}. Flow control refers to controlling the flow of execution of a
program, and branches of this type are loops and other control statements.
The second type is a \n{comparative branch}: these are comparisons between two
pieces of data, such as two keys in an array being sorted. A branch can be both
a flow control and a comparative branch, but these will be referred to as
comparative branches here. This is due to the properties of the branches: a
comparative branch should not be easily predictable; a flow control branch
should be easy to predict based on previous predictions of the same branch.

There are several kinds of branch predictors: \n{static}, \n{semi-static} and
\n{dynamic}. This report only deals with dynamic predictors, which are of
interest due to their high accuracy. The other types are none-the-less discussed
below, as an introduction to dynamic predictors. A good discussion on predictor
types is \cite{Uht97}, from which the percentages below are taken. These
predictors were tested by Uht on the SPECint92 benchmarking suite. This is a
standard suite of integer-based programs used to measure the performance of an
architecture against another. Programs in this benchmark include \cc{gcc}, a
major C compiler, \cc{espresso}, which generates Programmable Logic Arrays and
\cc{li}, a LISP interpreter.

### 2005-05-31 paper/docs/heap.tex
mostly removed the use of the word project, and 'I'

\section{Results}
All simulations in this project are performed on 32 byte cache lines. As a
result, 8-heaps are used for heapsort, since 8 4-byte \n{int}s fit the cache line.
LaMarca used 64 bit ints, and while this doesn't affect results, it means that
fewer keys fit into his cache lines. As a result, he used a 4-heap. To try and
emulate his results, extra tests were done, using a 4-heap with a 32 byte
cache line, and using a 4-heap using a 16 byte cache line, so that the entire
cache line is used. \verb!UINT_MAX! was used as the sentinel.

>>>>

\section{Results}
All simulations presented here were performed on 32 byte cache lines. As a
result, 8-heaps were used for heapsort, since 8 4-byte \n{int}s fit the cache line.
LaMarca used 64 bit ints, and while this doesn't affect results, it means that
fewer keys fit into his cache lines. As a result, he used a 4-heap. To try and
emulate his results, extra tests were done using a 4-heap with a 32 byte
cache line, and using a 4-heap using a 16 byte cache line, so that the entire
cache line is used. \verb!UINT_MAX! was used as the sentinel.
